---
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

# Appendix B: Point Emission Data {-}

## Overview

In this tutorial, we will demonstrate how to represent PM2.5 point emission data on maps. As an example, we will look at the PM2.5 data for four states - Illinois, Indiana, Wisconsin, and Michigan. The goal of this exercise is to create a point data map that provides a clear visualization of the point source of PM2.5 emission. To summarize, our objectives are to:

* Gain famililarity with the pollution data from 2014 National Emissions Inventory
* Perform simple data manipulation on the PM 2.5 data
* Visualize PM 2.5 pollution using the "tmap" package

## Environment Setup

### Input/Output {-}
The files that will be used in this tutorial are the pollution data for Illinois, Indiana, Wisconsin, and Michigan as well as he shapefile of the four states. The files are available for download on this project's [GitHub repo](https://github.com/GeoDaCenter/OpenAirQ-toolkit/tree/master/data).

### Load Libraries {-}

We start by loading the necessary packages - `tidyverse`, `sf`, and `tmap`: 

* `tidyverse`: to conduct basic statistical analyses
* `sf`: to perform simple spatial data manipulation. 
* `tamp`: to create spatial data visualization

```{r B_package.setup.points, message = FALSE}
library(tidyverse)
library(sf)
library(tmap)
```

### Load Data {-}

Besides the packages, we also need to load our data, which can be done by running the commented-out code below. We call the data frame `pe1`. 

```{r eval = FALSE, warning = FALSE, message = FALSE}
pe1 <- read_csv("process_12345.csv")
```


Note that this file above is unfortunately too large to be uploaded to Github, so we will instead use a pre-processed data set:


```{r echo = FALSE, warning = FALSE, message = FALSE}
fourstate.pe <- read_csv("./data/four_state.csv")
```


We also load the shapefile for the four states:


```{r}
fourstates <- st_read("./data/FourStates")
```

For more detailed description of the data, please refer to the main chapters of this tutorial book. **Insert Link!**

## Data Manipulation

Once we have our data and the packages ready, we will start the data manipulation process. Since we will only look at the data from Illinois, Indiana, Wisconsin, and Michigan, we can use the `filter` function to pick out only the four states that we are interested in, and we name the new data frame `fourstate.pe1`. This is accomplished by the commented-out code below.

This dataset comes from 2014 National Emissions Inventory, which is "a comprehensive and detailed estimate of air emissions of criteria pollutants, criteria precursors, and hazardous air pollutants from air emissions sources." You can read more about this dataset on this [website](https://www.epa.gov/air-emissions-inventories/national-emissions-inventory-nei).

```{r B_filtering_states, eval = FALSE}
states.abbr <- c("IL", "IN", "WI", "MI")

fourstate.pe <- pe1 %>%
        filter(state %in% states.abbr)
```

(Remark: The lines of code above does not need to be run if the data file loaded is `four_state.csv`. They were written to clean the data from `process_12345.csv`.)

Next, we turn our focus to the pollutant data. To familiarize ourselves with the variable `pollutant desc`, we use the `unique` function to examine all the unique values of this variable. 

```{r B_inspecting PM 2.5}
#Find pollutant names for pm2.5
unique(fourstate.pe$`pollutant desc`)
```

We see that there are quite a number of different pollutants, but we are primarily interested in the PM2.5 data. Therefore, we use the `filter` function again to subset the data, retaining only those observations with the pollutant being "PM2.5 Filterable" or "PM2.5 Primary (Filt + Cond)".

```{r C_filtering PM 2.5}
pm25 <- c("PM2.5 Filterable",  "PM2.5 Primary (Filt + Cond)")

#Filter for pm2.5
fourstate.pm <- fourstate.pe %>%
        filter(`pollutant desc` %in% pm25)
```

Now we will take care of the duplicates and missing values, both of which should be removed from our data. The following line of code eliminates any duplicated values in `eis facility id`.

```{r C_eliminating duplicates}
#remove duplicates
fourstate.pm.final <- fourstate.pm[!duplicated(fourstate.pm$`eis facility id`),] 
```

The following line of code eliminates any missing values in `site latitude`. We call this cleaned data frame `fourstate.pm.final`.

```{r C_eliminating missing values}
#remove na coords 
fourstate.pm.final <- fourstate.pm.final[!is.na(fourstate.pm.final$`site latitude`),] 
```

The last step in the data manipulation process is to turn our data points into a spatial object, which we can accomplish by using the `st_as_sf` function. Notice that we use the coordinate reference system 4326, which is the geodetic coordinate system for world. More information about coordinate reference systems can be found on this [website](https://epsg.io).

```{r C_making spatial object}
#Turn into sf object
fourstate.pm.spatial <- st_as_sf(fourstate.pm.final, coords = c("site longitude", "site latitude"), crs = 4326)
```

## Making Point Data Maps

```{r echo=FALSE, message = FALSE, warning = FALSE}
tmap_mode("view")
```

Finally, we are ready to make some maps! The command for generating a point data map is actually quite easy. We just need to specify the shapefile that stores the shape of the states and the point data which we have turned into a spatial object earlier. We adjust the size of the dots to 0.01 so that the pattern of those points is discernible. 

```{r C_making a map - tm_dots}
tm_shape(fourstates) +
        tm_borders() +
        tm_shape(fourstate.pm.spatial) +
        tm_dots(size = 0.01) 
```

The map above is neat, but it only conveys limited information. It is impossible, for example, to tell from the map how much PM2.5 emission each of these dots produces. 

To improve the simple map above, we use the `tem_bubbles` function instead of `tm_dots`. Within `tm_bubbles`, we can choose how to classify the PM2.5 data by specifying the `style`. Some common choices of `style` include "fisher", "jenks", "quantile", etc. Customization of the color palette is also possible, and there are plenty of online tutorials on this topic. 

```{r C_making a map - tm_bubbles}
tm_shape(fourstates) +
        tm_borders() +
        tm_shape(fourstate.pm.spatial) +
        tm_bubbles(col = "total emissions", size = 0.01,
                   style = "fisher", palette = "Reds") 
```

As we can see, the map above is not only more aesthetically pleasing, but it also communicates more information regarding the quantity of PM2.5 that is produced by each site. Sites that generate more PM2.5 are shown in a darker red. With this map, we can identify those sites of heavy PM2.5 emission with great ease by zooming in.

This concludes our tutorial. Following the simple steps above would allow you to create some simple point data maps, which are often a neat and easily interpretable visualization of the spatial data that we seek to analyze. 