---
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---
# Appendix I: Land Use Data {-}

## Overview

In this tutorial, we will take a close look at the land use layer data of those counties around Chicago, and our goal is to map out the cover percentage of areas that are covered by different categories of land use. At this point, it may not be clear what it is exactly that we intend to do, but everything will start to make sense once we inspect the available data. To be clear, our objective are to:

* Conduct data manipulation using "for loops"
* Calculate percentages for each grid
* Write new variables into the existing grid data file

## Environment Setup

### Load Libraries {-}

We will use the following packages in this tutorial:

* `raster`: to manipulate and analyze raster data
* `sf`: to conduct basic spatial data manipulation
* `tamp`: to create spatial data visualization
* `leadlet`: to create modern data visualization

```{r A_package.setup, message = FALSE}
library(rgdal)
library(raster)
library(tmap)
library(rgeos)
library(sp)
library(leaflet)
```

### Load Data {-}

We firstly load the data layers, set the projection, and name all land use categories.

```{r}
all = readOGR(dsn = ".", layer = "all")

CRS.new <- CRS("+init=EPSG:4326")
proj = spTransform(all, CRS.new)
crs(proj)

unique(all@data$LANDUSE)
```

## Data Manipulation

We firstly create filters for our data categories to convert the land use from numerical expressions to categories.

```{r}
resids = as.numeric(proj@data$LANDUSE) <= 5
coms = as.logical((as.numeric(proj@data$LANDUSE) >= 6) * (as.numeric(proj@data$LANDUSE) <= 13))
insts = as.logical((as.numeric(proj@data$LANDUSE) >= 14) * (as.numeric(proj@data$LANDUSE) <= 22))
inds = as.logical((as.numeric(proj@data$LANDUSE) >= 23) * (as.numeric(proj@data$LANDUSE) <= 28))
tcuws = as.logical((as.numeric(proj@data$LANDUSE) >= 29) * (as.numeric(proj@data$LANDUSE) <= 40) + (as.numeric(proj@data$LANDUSE) == 58))
ags = as.numeric(proj@data$LANDUSE) == 41
opens = as.logical((as.numeric(proj@data$LANDUSE) >= 42) * (as.numeric(proj@data$LANDUSE) <= 46) + (as.numeric(proj@data$LANDUSE) == 56))
vacs = as.logical((as.numeric(proj@data$LANDUSE) >= 47) * (as.numeric(proj@data$LANDUSE) <= 54))
waters = as.logical((as.numeric(proj@data$LANDUSE) == 55) + (as.numeric(proj@data$LANDUSE) == 57))
invs = as.numeric(proj@data$LANDUSE) >= 59
```

Then we save each category as a subset.

```{r}
residential = subset(proj, resids)
commercial = subset(proj, coms)
institutional = subset(proj, insts)
industrial = subset(proj, inds)
tcuw = subset(proj, tcuws)
agriculture = subset(proj, ags)
open_space = subset(proj, opens)
vacant_construction = subset(proj, vacs)
water = subset(proj, waters)
invalid = subset(proj, invs)
```

Now we save each category of land use as a new set of shapefiles.

```{r}
writeOGR(proj, dsn = ".", layer = "proj", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(residential, dsn = ".", layer = "residential", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(commercial, dsn = ".", layer = "commercial", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(institutional, dsn = ".", layer = "institutional", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(industrial, dsn = ".", layer = "industrial", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(tcuw, dsn = ".", layer = "tcuw", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(agriculture, dsn = ".", layer = "agriculture", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(open_space, dsn = ".", layer = "openspace", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(vacant_construction, dsn = ".", layer = "vacantconstruction", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(water, dsn = ".", layer = "water", driver = "ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(invalid, dsn = ".", layer = "invalid", driver = "ESRI Shapefile", overwrite_layer = TRUE)
```

## Load Community Shapefile

We now load the shapefiles for the community areas in Chicago.

```{r}
comarea = readOGR(dsn = ".", layer = "geo_export_c6dd6fe7-da82-4f79-b23b-ef61336a7389")
new_crs = "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

comarea = spTransform(comarea, new_crs)
keep = c("LANDUSE", "ShapeLen", "Shape_Area", "community", "area_numbe")
```

## Combine data files

Now we combine the land use shapefiles with the community areas shapefiles, and we save the results as new shapefiles for each land use category. The command for this process is `intersect`.

```{r}
residential = readOGR(dsn = ".", layer = "residential")
residential = intersect(residential, comarea)
residential@data = residential@data[keep]
writeOGR(residential, dsn = ".", layer = "residential_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

commercial = readOGR(dsn = ".", layer = "commercial")
commercial = intersect(commercial, comarea)
commercial@data = commercial@data[keep]
writeOGR(commercial, dsn = ".", layer = "commercial_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

institutional = readOGR(dsn = ".", layer = "institutional")
institutional = intersect(institutional, comarea)
institutional@data = institutional@data[keep]
writeOGR(institutional, dsn = ".", layer = "institutional_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

industrial = readOGR(dsn = ".", layer = "industrial")
industrial = intersect(industrial, comarea)
industrial@data = industrial@data[keep]
writeOGR(industrial, dsn = ".", layer = "industrial_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

tcuw = readOGR(dsn = ".", layer = "tcuw")
tcuw = intersect(tcuw, comarea)
tcuw@data = tcuw@data[keep]
writeOGR(tcuw, dsn = ".", layer = "tcuw_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

agriculture = readOGR(dsn = ".", layer = "agriculture")
agriculture = intersect(agriculture, comarea)
agriculture@data = agriculture@data[keep]
writeOGR(agriculture, dsn = ".", layer = "agriculture_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

open_space = readOGR(dsn = ".", layer = "openspace")
open_space = intersect(open_space, comarea)
open_space@data = open_space@data[keep]
writeOGR(open_space, dsn = ".", layer = "open_space_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

vacant_construction = readOGR(dsn = ".", layer = "vacantconstruction")
vacant_construction = intersect(vacant_construction, comarea)
vacant_construction@data = vacant_construction@data[keep]
writeOGR(vacant_construction, dsn = ".", layer = "vacant_construction_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

water = readOGR(dsn = ".", layer = "water")
water = intersect(water, comarea)
water@data = water@data[keep]
writeOGR(water, dsn = ".", layer = "water_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)

invalid = readOGR(dsn = ".", layer = "invalid")
invalid = intersect(invalid, comarea)
invalid@data = invalid@data[keep]
writeOGR(invalid, dsn = ".", layer = "invalid_chi", driver = "ESRI Shapefile", overwrite_layer = TRUE)
```

## Calculate Land Use Coverage

We firstly create a function `coverage` to calculate the coverage percentage for each land use category of the total area of each community area.

```{r}
coverage = data.frame(residential = rep(0,nrow(comarea@data)), commercial = rep(0,nrow(comarea@data)), institutional = rep(0,nrow(comarea@data)),industrial = rep(0,nrow(comarea@data)), tcuw = rep(0,nrow(comarea@data)), agriculture = rep(0,nrow(comarea@data)),open_space = rep(0,nrow(comarea@data)), vacant_construction = rep(0,nrow(comarea@data)), water = rep(0,nrow(comarea@data)),invalid = rep(0,nrow(comarea@data)), total = rep(0,nrow(comarea@data)))

rownames(coverage) = comarea@data$community
```

### Load New Data {-}

Now we load the shapefiles we got from the last step and get the total area of each community.

```{r}
residential_chi = readOGR(dsn = ".", layer = "residential_chi")

commercial_chi = readOGR(dsn = ".", layer = "commercial_chi")

institutional_chi = readOGR(dsn = ".", layer = "institutional_chi")

industrial_chi = readOGR(dsn = ".", layer = "industrial_chi")

tcuw_chi = readOGR(dsn = ".", layer = "tcuw_chi")

agriculture_chi = readOGR(dsn = ".", layer = "agriculture_chi")

open_space_chi = readOGR(dsn = ".", layer = "open_space_chi")

vacant_construction_chi = readOGR(dsn = ".", layer = "vacant_construction_chi")

water_chi = readOGR(dsn = ".", layer = "water_chi")

invalid_chi = readOGR(dsn = ".", layer = "invalid_chi")

total_area = comarea@data$shape_area
```

### Calculate Coverage Rate

We use the following for-loop to calculate the percentage for each land use category.

```{r}
for (i in 1:nrow(coverage)) {
  com_name = as.character(comarea@data$community[i])
  coverage[com_name, "residential"] = sum(residential_chi@data$Shape_Area * (residential_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "commercial"] = sum(commercial_chi@data$Shape_Area * (commercial_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "institutional"] = sum(institutional_chi@data$Shape_Area * (institutional_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "industrial"] = sum(industrial_chi@data$Shape_Area * (industrial_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "tcuw"] = sum(tcuw_chi@data$Shape_Area * (tcuw_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "agriculture"] = sum(agriculture_chi@data$Shape_Area * (agriculture_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "open_space"] = sum(open_space_chi@data$Shape_Area * (open_space_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "vacant_construction"] = sum(vacant_construction_chi@data$Shape_Area * (vacant_construction_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "water"] = sum(water_chi@data$Shape_Area * (water_chi@data$community == com_name)) / total_area[i]
  coverage[com_name, "invalid"] = sum(invalid_chi@data$Shape_Area * (invalid_chi@data$community == com_name)) / total_area[i]
}

comarea@data$residential = coverage[, "residential"]
comarea@data$commercial = coverage[, "commercial"]
comarea@data$institutional = coverage[, "institutional"]
comarea@data$industrial = coverage[, "industrial"]
comarea@data$tcuw = coverage[, "tcuw"]
comarea@data$agriculture = coverage[, "agriculture"]
comarea@data$open_space = coverage[, "open_space"]
comarea@data$vacant_construction = coverage[, "vacant_construction"]
comarea@data$water = coverage[, "water"]
comarea@data$invalid = coverage[, "invalid"]
```

### Write the Results

We write the results as a new set of shapefiles as the following: 

```{r}
writeOGR(comarea, dsn = ".", layer = "comarea_ratios", driver = "ESRI Shapefile")
```

