---
title: "Temperature Tutorial"
author: "Yuming Liu"
date: "October 20, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the Packages

We can use the "Packages" set-up on the right hand side to install the packages, or we could just run the code `install.packages('name', dependencies=TRUE, repos='http://cran.rstudio.com/')` from the Console to install new packages.

Spatial analysis in R requires multiple libraries. Package installation is done with the following syntax: `install.packages(“sp”)`. Some of these take additional time for installation, depending on your system. The following list is comprehensive for this tutorial, as well as much spatial analysis.

```{r packages,include=FALSE}
library(sp) #spatial data wrangling & analysis
library(rgdal) #spatial data wrangling & analysis
library(rgeos) #spatial data wrangling & analysis
library(leaflet) #modern data visualizations
library(raster) #spatial data wrangling & analysis
library(gstat) #spatial data wrangling & analysis
library(tmap) #modern data visualizations
library(tidyverse) # data wrangling
```

## Import Data

We can read the monthly NOAA data using the following lines:

```{r data}
noaa.monthly <- read.csv("NOAA_master_monthly_final.csv")
glimpse(noaa.monthly)
```

## Data Wrangling

We use the following code to select minimum monthly temperatures from the data:

```{r min}
temp.monthly.min <- noaa.monthly %>%
  dplyr::select(contains("min_mo"))
```

Then, for example, we isolate the data for Summer 2015.

```{r summer}
temp.monthly.min <- noaa.monthly %>%
  dplyr::select(contains("min_mo"))
```

We get the average minumum low temperature over the summer.

```{r meanlow}
sum.mintemp.2015 <- temp.monthly.min %>%
  dplyr:: select(`X2015_6_temp_min_mo`:`X2015_8_temp_min_mo`)

sum.mintemp.2015 <- rowSums(sum.mintemp.2015, na.rm = TRUE)
```

## Merge Temperature data to Sensors

We use the following code to select sensor coordinates and other information. 

```{r sensor}
sensor.info <- noaa.monthly %>%
  dplyr::select(STATION:elevation)

#Combine the sensor info with the sensor readings
sensor.temp.min <- cbind(sensor.info, sum.mintemp.2015)
tail(sensor.temp.min)

#Remove the 0s because 0 from the previous line, the NAs are now 0s
sensor.temp.min[sensor.temp.min==0] <- NA
complete.temp.min <- na.omit(sensor.temp.min)

glimpse(complete.temp.min)
```

## Ploting the Results

Now we set the longtitude and latitude and the projection.

```{r lon/lat}
#Set lat/lon
coordinates(complete.temp.min) <- complete.temp.min[,c("long", "lat")]
#Set projection to WSG84
proj4string(complete.temp.min) <- CRS("+init=epsg:4326")
```

Then we import the map of Chicago Community Areas.

```{r}
chi.map <- readOGR("Chicago")

chi.map <- spTransform(chi.map, CRS("+init=epsg:4326 +proj=longlat +datum=WGS84 +no_defs +ellps=WGS84
                                    +towgs84=0,0,0"))
```

## Creating a Variogram

A variogram would be used to show the variance between different sensors. The variogram is defined as the variance of the difference between field values at two locations ({\displaystyle \mathbf {s} _{1}}{\mathbf  {s}}_{1} and {\displaystyle \mathbf {s} _{2}}{\displaystyle \mathbf {s} _{2}}, note change of notation from {\displaystyle M}M to {\displaystyle \mathbf {s} }\mathbf {s}  and {\displaystyle f}f to {\displaystyle Z}Z) across realizations of the field.

```{r}
sum.mintmp.vgm <- variogram(complete.temp.min$sum.mintemp.2015 ~ 1, complete.temp.min)
plot(sum.mintmp.vgm)
```

```{r}
sum.fit <- fit.variogram(sum.mintmp.vgm, model=vgm("Sph"))
plot(sum.mintmp.vgm, sum.fit)
```

## Visualization

We firstly generate a prediction surface grid. Then we fit the grid with the map of Chicago.

```{r}
pt2grid <- function(ptframe,n) {
  bb <- bbox(ptframe)  
  ptcrs <- proj4string(ptframe)  
  xrange <- abs(bb[1,1] - bb[1,2])  
  yrange <- abs(bb[2,1] - bb[2,2])  
  cs <- c(xrange/n,yrange/n)  
  cc <- bb[,1] + (cs/2)  
  dc <- c(n,n)  
  x1 <- GridTopology(cellcentre.offset=cc,cellsize=cs,cells.dim=dc)  
  x2 <- SpatialGrid(grid=x1,proj4string=CRS(ptcrs))
  return(x2)
}

chi.grid <- pt2grid((chi.map),100)
```

In statistics, originally in geostatistics, kriging or Gaussian process regression is a method of interpolation for which the interpolated values are modeled by a Gaussian process governed by prior covariances. In this case, kriging gives the best linear unbiased prediction of the intermediate values.

```{r}
sum.kriged <- krige(complete.temp.min$sum.mintemp.2015 ~ 1, complete.temp.min, chi.grid, model = sum.fit)
plot(sum.kriged)
```

Then we finish the visualization by ploting the average low temperature for Chicago.
```{r}
chi.sum.kriged <- sum.kriged[chi.map,]
plot(chi.sum.kriged)
title(main = "Average Low Temperature (F) Summer 2015", outer = FALSE)
```
